R for Clinical Study Reports and Submission

### 2021-10-04 project-folder.Rmd

In this book, we refer an R package as **project specific R package**. 
The purpose of project specific R package is to organize analysis scripts for a clinical project.
We refer an R package as **standard R package** if the purpose of an R package is to 
share commonly used R functions to be hosted in an code repository such as CRAN.

Below is a minimal sufficient folders and files based on the R package folder structure.

@@

To illustrate the idea, you can open `esubdemo` project by click the `esubdemo.Rproj` project.

For analysis code to create TLFs, we saved them in the `vignettes/` folder. 
In `esubdemo`, we put analysis code we learned from previous chapters into the `vignettes` folder.

@@

In previous chapters, we also create a few helper functions that can be used for different analysis. 
(e.g. `fmt_num` and `count_by`). 
We put those helper functions into the `R/` folder.

> Here `zzz.R` is a template file that ensure startup file `inst/startup.R` is executed. 
> We will revisit startup file while we discuss reproducibility.

```
R
├── count_by.R
├── fmt.R
└── zzz.R
```

@@

Reproducibility of analysis can be challenge. 
By using R package folder structure and proper tools (e.g. `renv`, `packrat`), 
we illustrate how to achieve reproducibility for R and R package versions.

@@

In `esubdemo`, the `.Rprofile` contain two lines of R code.

```{r}
source("inst/startup.R")
source("renv/activate.R")
```

The first line execute R code saved in `inst/startup.R`. 
People can define a customized startup file to check running environment.

For example, we can require R version to be `4.1.1` by using code below. 
```{r, eval = FALSE}
R_version <- "4.1.1"	# set up project R version

# Check R Version
if(paste(R.version$major, R.version$minor, sep = ".") != R_version){
 warning("The current R version is not the same with the current project in ", R_version)
}
```

@@

We can also define the package repository to be a specific snapshot date.
For example, we used Microsoft R Application Network (MRAN) to define the snapshot date to be `2021-08-06`. 
The snapshot date freeze the R package repository. 
In other words, all R packages installed in this RStudio project 
are based on the frozen R version at the snapshot date. 
Here it is `2021-08-06` by using MRAN server.

@@

@@

The second line execute R code in `renv/activate.R`. 
This is a default behavior by using the `renv` R package to manage snapshot 
of the R packages. 
The `renv` provide a robust and stable approach to manage R package version 
for a project specific R packages.
As a user, you can use `renv::init()`, `renv::snapshot()` and `renv::restore()` 
to initialize, save and restore R packages used for the current analysis project.

@@

An introduction of `renv` can be found in its [package website](https://rstudio.github.io/renv/articles/renv.html).

@@

```
* Project '~/esubdemo' loaded. [renv 0.14.0]
Current oroject R package repository:
    https://mran.microsoft.com/snapshot/2021-08-06
```

@@

```
Below R package path are searching in order to find installed R pacakges in this R session:
    /home/esubdemo/renv/library/R-4.0/x86_64-pc-linux-gnu
    /rtmp/Rtmp8d5mcL/renv-system-library
```

@@

Here is a short list of commonly used R packages

- [`devtools`](https://devtools.r-lib.org/): make package development easier. 
  + A good overview can be found in [Chapter 2 of the R packages](https://r-pkgs.org/whole-game.html) book  
- [`usethis`](https://usethis.r-lib.org/): automates repetitive tasks that arise during project setup and development.
- [`testthat`](https://testthat.r-lib.org/): streamline testing code.
  + A discussion of using the `testthat` for an A&R project can be found in (@madhutest).  
- [`pkgdown`](https://pkgdown.r-lib.org/): generate static html documentation for an R package
  + It also allow you to run all analysis code in batch.

@@

- ToDo

>>>>

In this book, we refer an R package as **project specific R package** 
if the purpose of an R package is to organize analysis scripts for a clinical project.

We refer an R package as **standard R package** if the purpose of an R package is to 
share commonly used R functions to be hosted in an code repository such as CRAN.

Below is a minimal sufficient folders and files for a project specific R package
based on the R package folder structure.

@@

A general discussion of the R package folder structure can be found 
in Chapter 4 of the R package book (@wickham2015r).

We demonstrate the idea using the [`esubdemo`](https://github.com/elong0527/esubdemo) project.

In `esubdemo` project, we saved all TLF generation scripts in previous chapters into the `vignettes/` folder.

@@

While creating those analysis scripts, we also defined a few helper functions (e.g. `fmt_num` and `count_by`). 
Those functions are saved in the `R/` folder.

```
R/
├── count_by.R
└── fmt.R
```

With a formal project, it is also important to provide proper documentations for 
those help functions. We used `Roxygen2` to document functions. 
For example, the header below provide the definition of each variable in `fmt_est`. 
More details can be found in [Chpater 10 of the R package book](https://r-pkgs.org/man.html)

```{r, eval = FALSE}
#' Format point estimator
#'
#' @param .mean mean of an estimator.
#' @param .sd sd of an estimator.
#' @param digits number of digits for `.mean` and `.sd`.
#'
#' @export
fmt_est <- function(.mean,
                    .sd,
                    digits = c(1, 2)) {
  .mean <- fmt_num(.mean, digits[1], width = digits[1] + 4)
  .sd <- fmt_num(.sd, digits[2], width = digits[2] + 3)
  paste0(.mean, " (", .sd, ")")
}
```

Those `Roxygen2` documentation will be transferred to 
standard R documentation `.Rd` files saved in the `man/` folder. 
This step is automatically handled by `devtools::document()` function.

```
man
├── count_by.Rd
├── fmt_ci.Rd
├── fmt_est.Rd
├── fmt_num.Rd
└── fmt_pval.Rd
```

@@

Reproducibility of analysis is one of the most important aspects for regulatory deliverables. 
By using R package folder structure and proper tools (e.g. `renv`, `packrat`), 
we illustrate how to achieve reproducibility for R and R package versions.

@@

After we open the `esubdemo` project, the code in `.Rprofile` will automatically check 
current R version is the same as we defined in `.Rprofile`

```{r, eval = FALSE}
R_version <- "4.1.1" # set up project R version
```

@@

The R package version is controlled in two layers. 
Firstly, we define a snapshot date in `.Rprofile`. 
The snapshot date allows us to freeze the source code repository.

@@

We can also define the package repository to be a specific snapshot date.
For example, we used Microsoft R Application Network (MRAN) to 
define the snapshot date to be `2021-08-06`. 
The snapshot date freeze the R package repository.

In other words, all R packages installed in this RStudio project 
are based on the frozen R version at the snapshot date. 
Here it is `2021-08-06` by using MRAN server.

Below information will be displayed after a new R session is opened.

```
Current oroject R package repository:
    https://mran.microsoft.com/snapshot/2021-08-06
```

@@

Secondly, we use `renv` to lock R package versions and saved in `renv.lock` file. 
The `renv` provide a robust and stable approach to manage R package version 
for a project specific R packages.
An introduction of `renv` can be found in its 
[package website](https://rstudio.github.io/renv/articles/renv.html).

```{r}
source("renv/activate.R")
```

The R code above in the `.Rprofile` initiate the `renv` running environment.  
As a user, you can use `renv::init()`, `renv::snapshot()` and `renv::restore()` 
to initialize, save and restore R packages used for the current analysis project.

@@

@@

```
* Project '~/esubdemo' loaded. [renv 0.14.0]
```

@@

```
Below R package path are searching in order to find installed R pacakges in this R session:
    "/home/zhanyilo/github-repo/esubdemo/renv/library/R-4.1/x86_64-pc-linux-gnu"
    "/rtmp/RtmpT3ljoY/renv-system-library"   
```

@@

We have learned a few functions in `devtools` to automatically generate content. 
Here is a short list of tools that can enhance the workflow.

- [`devtools`](https://devtools.r-lib.org/): make package development easier. 
  + A good overview can be found in [Chapter 2 of the R packages](https://r-pkgs.org/whole-game.html) book  
  + `devttols::load_all()`: load all functions in `R/` folder and running environment. 
  + `devtools::document()`: automatically create documentation using `Roxygen2`
  + `devttols::check()`: automatically perform compliance check as an R package. 
  + `devtools::build_site()`: automatically run analysis scripts in batch and create a `pkgdown` website. 
- [`usethis`](https://usethis.r-lib.org/): automates repetitive tasks that arise during project setup and development.
- [`testthat`](https://testthat.r-lib.org/): streamline testing code.
  + A discussion of using the `testthat` for an A&R project can be found in (@madhutest).  
- [`pkgdown`](https://pkgdown.r-lib.org/): generate static html documentation for an R package
  + It also allow you to run all analysis code in batch.

You may further automatically execute routines by leveraging CI/CD workflow. 
For example, the `esubdemo` project will rerun all required check and build pkgdown website
by using [github action](https://usethis.r-lib.org/reference/github_actions.html).

As the consistent folder is defined, it also become easier to create specific tools 
that fit for the analysis and reporting purpose.

Below are few potential tools that can be helpful:

- Create project template using [RStudio Project Templates](https://rstudio.github.io/rstudio-extensions/rstudio_project_templates.html)
- Add additional compliance check for analysis and reporting. 
- Save log files for running in batch.

@@

For a regulatory deliverable, it is important to maintain compliance. 
With a consistent folder structure, we can define specific criteria for compliance. 
Some compliance criteria can be implemented within the automatically checking steps.

For an R package, there are already criteria to ensure R package integrity.  
More details can be found in [Chapter 19 of the R package book](https://r-pkgs.org/r-cmd-check.html).
